---
title: "Calibrating model parameters with the calibration package"
author:
  - "Akash BV"
format:
  html: default
  pdf: default
toc: true
number-sections: true
fig-width: 9
fig-height: 6
fig-dpi: 200
---

# Introduction

This notebook walks you through the `calibration` package: what it does, how it
is put together, and how to run it. By the end you will have run a small example
yourself and seen how the same steps drive a real SIPNET soil carbon calibration.

## What the package does

Calibration means finding the parameter values that make a process model agree
with field data, and saying how confident we are in those values. This package
does that with Ensemble Kalman Inversion (EKI): it runs an ensemble of parameter
sets through the model and nudges them toward the values that fit the
observations.

Two things to know up front:

- It is not tied to one model, site, or variable. The observations, the sites,
  the parameters, and the model all come in through arguments, so the same code
  calibrates soil carbon at one site or fluxes at many.
- It is set up so a new method can be added later. Right now the method is EKI;
  the entry point `calibrate()` dispatches to it, and other methods can slot in
  the same way.

## What this notebook does

1. Explains the four pieces you provide to a calibration.
2. Runs a small example that finishes in a few seconds, so you can see the loop
   work end to end.
3. Shows the same steps on the real SIPNET soil carbon run at Salinas.

## Prerequisites

Install the package from its source directory:

```{r install, eval = FALSE}
# from the parent of the calibration/ directory
install.packages("calibration", repos = NULL, type = "source")
# or, during development
pkgload::load_all("calibration")
```

The SIPNET section also needs PEcAn and the prepared forward run, which live on
the cluster. The small example needs only the package itself.

```{r libraries, message = FALSE}
library(calibration)
```

# The four pieces of a calibration

Every calibration is one call:

```{r contract, eval = FALSE}
result <- calibrate(obs, prior, forward, control)
```

and you provide four things:

- **`obs`** -- what you are fitting to. A named vector `y` of observed values,
  their covariance `Sigma` (how much you trust each one), and a `meta` table that
  keeps track of what each value is.
- **`prior`** -- your starting belief about the parameters, before seeing the
  data. One record per parameter with its distribution and its allowed range.
- **`forward`** -- a function that runs the model at a set of parameter values and
  returns what the model predicts at each observation. This is the only piece
  that knows about the model.
- **`control`** -- which method to use and its settings (ensemble size,
  iterations, seed), from `calibration_control()`.

`calibrate()` returns the posterior: the updated ensemble of parameters (`U`),
the starting ensemble (`U0`), and the per iteration model predictions you can
plot and score.

# A small example you can run now

Here the model is the identity: the prediction equals the parameter, so we know
the true answer and can check the calibration finds it. The prior starts near
zero and the observations are (2, -1), measured tightly, so the data should lead.

```{r toy-setup}
# prior: two parameters, each a wide normal centred at zero
prior <- prior_from_specs(list(
  theta1 = list(distn = "norm", parama = 0, paramb = 5),
  theta2 = list(distn = "norm", parama = 0, paramb = 5)
))

# observations: two values, measured tightly (small variance)
y <- c(obs1 = 2, obs2 = -1)
Sigma <- diag(c(0.05, 0.05))
dimnames(Sigma) <- list(names(y), names(y))
obs <- list(y = y, Sigma = Sigma)

# forward model: prediction equals the parameter
forward <- function(U, iteration) {
  G <- U
  colnames(G) <- names(y)
  G
}
```

Run it:

```{r toy-run}
result <- calibrate(
  obs, prior, forward,
  control = calibration_control(n_particles = 300, n_iterations = 3, seed = 1)
)

rbind(
  prior     = round(colMeans(result$U0), 2),
  posterior = round(colMeans(result$U), 2),
  target    = c(2, -1)
)
```

The posterior mean moves from near zero to the target. That is the whole loop:
start from the prior, run the model, pull toward the data, repeat.

You can see the shift per parameter:

```{r toy-densities, eval = FALSE}
plot_param_densities(
  list(prior = result$U0, posterior = result$U),
  pdf_file = "toy_param_densities.pdf"
)
```

# Calibrating SIPNET soil carbon

The same four pieces, now with real data and SIPNET as the model. The values here
are the Salinas organic cropping systems soil carbon run; the surrounding code is
the general package.

## The observations

`build_obs()` reads the curated cal/val data and builds the target for one
variable at the sites you name. Each site, treatment, and year becomes one slot;
the value is the replicate mean and the variance comes from the replicate spread.

```{r sipnet-obs, eval = FALSE}
obs <- build_obs(
  cal_val_dir = "/projectnb/dietzelab/ccmmf/usr/akash/cal_val",
  target_var  = "SOC_stock_Mg_ha",
  sites       = "salinas_socs"
)
```

## The prior

Priors come from three sources, combined into one list:

- traits with a meta-analysis posterior are read straight from it,
- traits without one get a plausible range written down explicitly,
- a calibrated initial state is anchored to its first observation.

```{r sipnet-prior, eval = FALSE}
prior <- c(
  prior_from_postdistns("som_respiration_rate", soil_pft$posterior.files),
  prior_from_specs(list(
    soil_respiration_Q10 = list(distn = "unif", parama = 1.4,  paramb = 3.0),
    turn_over_time       = list(distn = "unif", parama = 0.13, paramb = 1.2)
  )),
  state_prior_from_obs(obs$meta, prefix = "soilInit.",
                       from_unit = "Mg/ha", to_unit = "kg/m2", anchor_year = 2)
)
```

## The forward model

`make_forward_sipnet()` wraps a prepared PEcAn multisite run as the `forward`
function. It runs the ensemble through SIPNET, using the settings and the SGE
launcher exactly as prepared, and harvests the variable back to the observation
units.

```{r sipnet-forward, eval = FALSE}
settings <- PEcAn.settings::read.settings(
  "/projectnb/dietzelab/ccmmf/usr/akash/salinas_socs/settings.xml"
)

forward <- make_forward_sipnet(
  settings = settings, obs = obs, n_particles = 50,
  harvest_var = "TotSoilCarb", from_unit = "kg/m2", to_unit = "Mg/ha",
  soil_pft = "soil", state_prefix = "soilInit.",
  state_pool = "soil_organic_carbon_content"
)
```

## Run it

This launches the SIPNET ensemble on the cluster, so it is not run here. In
practice you run it through the numbered scripts (`scripts/010` to `040`) against
a config; see `examples/salinas_soc/`.

```{r sipnet-run, eval = FALSE}
result <- calibrate(
  obs, prior, forward,
  control = calibration_control(n_particles = 50, n_iterations = 3, seed = 556688)
)
```

## Look at the results

The result carries the per iteration model predictions and the prior and
posterior parameter ensembles, which the plotting and scoring functions consume.

```{r sipnet-results, eval = FALSE}
G_list <- c(lapply(result$eki$eki_list, function(s) s$G), list(result$eki$G))
names(G_list) <- paste("iteration", seq_along(G_list))

# fit of each iteration against the data
score_table(G_list, result$obs$meta)

# ensemble against the observations, one page per iteration
save_iterations_pdf(G_list, result$obs$meta, "ensembles_vs_truth.pdf",
                    ylab = "SOC stock (Mg C ha-1)")

# how each parameter moved from prior to posterior
plot_param_densities(list(prior = result$U0, posterior = result$U),
                     "param_prior_posterior.pdf")
```

The score table reports RMSE, bias, coverage, interval width, and CRPS per
iteration. The two figures show the ensemble tracking the observations and the
prior to posterior shift in each parameter.

# Using it on your own problem

To calibrate something else, you change the four pieces, not the estimator:

- point `build_obs()` at a different `target_var` and `sites`,
- build a prior for the parameters you want,
- give `make_forward_sipnet()` (or your own `forward`) the variable and units,
- keep `calibrate()` as is.

The estimator only ever sees `y`, `Sigma`, and the model predictions, so it does
not care what the variable, the site, or the model are.

# Conclusion

You have run a calibration end to end on a toy problem and seen the same steps
set up the real SIPNET soil carbon run. The pattern is always the same: describe
the observations, state the prior, wrap the model, and call `calibrate()`.

# Session info

```{r session-info}
sessionInfo()
```
