---
title: "Statewide treatment-effect matrix: run report"
date: today
format:
  html:
    self-contained: true
    toc: true
    theme: cosmo
    number-sections: true
execute:
  echo: false
  warning: false
  message: false
---

```{r}
#| label: setup
#| include: false
library(dplyr)
library(ggplot2)
library(knitr)

config <- config::get(file = file.path(here::here(), "examples/3_statewide_effects/config.yml"))
ws <- config$workspace
score <- read.csv(file.path(ws, "model_vs_evidence_scorecard.csv"))
eff <- read.csv(file.path(ws, "site_pair_effects.csv"),
                colClasses = c(site_id = "character"))
# runs_manifest.csv is written by the PEcAn config step and lists every run it
# created, so the design minus the manifest is what never reached the model
manifest <- read.csv(file.path(ws, "output", "runs_manifest.csv"))
design <- read.csv(file.path(config$event_package$root, "run_pairs.csv"),
                   colClasses = c(site_id = "character"))

active <- score |> filter(target_use %in% c("likelihood", "sign_check"))
n_pairs <- nrow(eff)
n_sites <- dplyr::n_distinct(eff$site_id)
n_blocked <- nrow(dplyr::distinct(
  tidyr::pivot_longer(design, c(treatment, reference), values_to = "arm"),
  site_id, arm)) - nrow(manifest)
ef <- score |> filter(pair_id == "minN_EF_level")
ef_level <- ef$model_effect
ef_target <- ef$target_center

theme_set(theme_minimal(base_size = 12) +
  theme(panel.grid.minor = element_blank(), legend.position = "bottom"))
```

This report interprets one executed forward matrix of SIPNET practice effects
at the statewide design points, run with the configuration in this directory.
Nothing here reruns the model: every number is read from the run outputs under
`` `r ws` ``.

# What was run

`r n_pairs` treatment pairs at `r n_sites` sites, one run per (site, arm),
over `r config$window$start` to `r config$window$end`. Within a pair only
`events.in` differs; met drivers, initial conditions, parameters, executable
and configuration are shared. `r n_blocked` arms at one site are blocked on an
invalid wood initial condition and were not substituted.

Vegetation parameters are the plant functional type medians from the pinned
input package; photosynthesis optimum, soil respiration rates, soil C:N and
the nitrogen settings are run level and pinned in `config.yml`.

# Model against evidence

Each cell is scored with the operator recorded in the target table, and the
model side is aggregated to one estimate per cell before any comparison.

```{r}
#| label: scorecard
active |>
  transmute(
    cell = paste0(target_cell,
                  ifelse(is.na(target_subclass) | target_subclass == "", "",
                         paste0(" [", target_subclass, "]"))),
    use = target_use,
    target = signif(target_center, 4),
    spread = signif(target_spread, 3),
    spread_type = target_spread_type,
    model = signif(model_effect, 3),
    z = round(z_vs_target_spread, 2),
    n = model_n_sites,
    status = status
  ) |>
  kable()
```

`z` is the model estimate minus the target centre in units of the target's own
spread, and must be read with `spread_type`. A population standard deviation
asks whether the model sits inside the observed heterogeneity; a standard
error of a synthesis mean asks whether it matches that mean, and is tight
enough that any model value is many multiples away.

```{r}
#| label: site-effects
#| fig-height: 3.2
#| fig-width: 7
fam <- bind_rows(
  tibble(family = "cover, per cover yr",
         value = eff$dd_soc_per_cover_yr[eff$treatment == "cover_annual"]),
  tibble(family = "amendment, soil LRR",
         value = eff$soil_lrr_end[eff$treatment == "organic_substitution"]),
  tibble(family = "zero till, per yr",
         value = eff$dd_soc_per_window_yr[eff$treatment == "tillage_zero"]),
  tibble(family = "zero till, N2O LRR",
         value = eff$lrr_n2o_till[eff$treatment == "tillage_zero"])
) |>
  filter(is.finite(value))

ggplot(fam, aes(x = value, y = family)) +
  geom_vline(xintercept = 0, linewidth = 0.3, colour = "grey60") +
  geom_boxplot(outlier.size = 0.6, linewidth = 0.3, width = 0.5) +
  labs(x = "site-level paired effect", y = NULL)
```

Site-level effects behind four of the scored cells. The aggregate that enters
the scorecard is the mean of each distribution, not the individual sites.

# What the matrix shows

The three soil carbon rate cells agree with their targets in sign and sit
inside their spreads. Cover is positive at
`r round(100 * score$model_frac_positive[score$pair_id == "cover_soc"])` percent
of sites, reduced and zero tillage at 100 percent, and the amendment cell at
100 percent.

Three cells cannot be matched by any parameter setting in the current model,
for reasons in the model structure rather than the parameterization.

Tillage enters as a heterotrophic respiration multiplier with no compaction or
pore-space pathway, so tillage nitrous oxide is a small negative log ratio at
every site. The cell is matched to the under-10-year adoption class, which
reports a positive effect; van Kessel finds the response flips with duration
and reports a negative effect at 10 years or more, so the modelled sign agrees
with the long-duration class rather than the one matched here.

The anaerobic index clips soil water at saturation, and the prescribed flood
keeps both arms above field capacity through the drying windows at eleven of
twelve rice sites. The scored quantity is the flooded-season total, and the two
arms give an identical season total at nine of twelve, so the rice methane
cells carry no model response rather than a small one.

Nitrous oxide is first order in the mineral nitrogen pool, so induced
emissions are proportional to nitrogen applied and the emission factor is
constant across the fertilized arms. The observed rise of the emission factor
with nitrogen rate cannot be reproduced. The level of that emission factor is
`r round(ef_level, 2)` percent of nitrogen applied against a Mediterranean
synthesis value of `r sprintf("%.2f", ef_target)` percent, but that cell is
reported without a discrepancy: the modelled level is an upper bound on the
measured quantity, and a bound above the target settles nothing.

# Limitations

The `n2o` output is total mineral nitrogen volatilization rather than a
partitioned nitrous oxide flux, so the emission factor level is an upper bound
on the measured quantity and its cell carries no discrepancy. The dose response
is a ratio of that same quantity, and is unaffected only if the nitrous oxide
share of volatilization does not itself move with nitrogen rate; if that share
varies with rate, a real dose response could exist while the quantity computed
here stays flat. The induced emission factor has also not equilibrated over the
window, rising monotonically without a plateau.

Two cells receive more than one row, since tillage soil carbon and rice drying
each map two arms to one cell; those rows are not independent estimates.
Cover results are sensitive to one site with a single prepared cover year on an
organic soil. The design points predate a later downscaling release that
swapped two parcels. For the fertilizer cell, the published crop classes match
this panel poorly, as most fertilized sites are orchard or pasture.

# Reproducing

The run is determined by this repository at the staging commit and by
`config.yml`, which pins the event package, the input paths, the model
revision and every parameter override. The script sequence is in the README.
