---
title: "Joint SOC and N2O calibration: run report"
date: today
format:
  html:
    self-contained: true
    toc: true
    theme: cosmo
    number-sections: true
execute:
  echo: false
  warning: false
  message: false
---

```{r}
#| label: setup
#| include: false
library(calibration)
library(dplyr)
library(ggplot2)
library(knitr)

config <- config::get(file = file.path(here::here(), "examples/2_joint_soc_n2o/config.yml"))
cache  <- config$cache_dir
result <- readRDS(file.path(cache, "result.rds"))
target <- readRDS(file.path(cache, "target.rds"))
obs    <- readRDS(file.path(cache, "obs.rds"))
prior  <- readRDS(file.path(cache, "prior.rds"))

G_list <- c(lapply(result$eki$eki_list, function(s) s$G), list(result$eki$G))
names(G_list) <- paste("iteration", seq_along(G_list))
shift <- param_shift(result$U0, result$U)
scores <- score_table(G_list, result$obs$meta)

theme_set(theme_minimal(base_size = 12) +
  theme(panel.grid.minor = element_blank(), legend.position = "bottom"))
```

This report interprets one executed calibration of SIPNET soil parameters
against the curated cal/val observations, run with the configuration in this
directory. The run itself is reproduced by the script sequence in the README;
nothing here reruns the model: every figure and number is read from the run's
cached artifacts under `` `r cache` ``.

# What was calibrated

The fitted target is soil organic carbon at the Salinas organic cropping
systems trial, contracted from `r sum(obs$meta$variable == "SOC_stock_salinas")`
per system per year stocks (8 systems, 2005 to 2011) into
`r sum(result$obs$meta$variable == "SOC_periodmean_salinas")` slots: the control
system's period mean level and one period mean contrast per remaining system.
The Modesto almond N2O treatment contrast
(`r sum(target$meta$variable == "N2O_contrast_modesto")` date-matched slots) is
held out of the likelihood and scored as validation.

Three soil parameters are estimated (`som_respiration_rate`,
`fracLitterRespired`, `turn_over_time`) together with one initial soil carbon
stock per system (`soilInit.*`), each anchored on that system's 2005 observation
at its measurement error rather than pinned to it. `soil_respiration_Q10` is
pinned at 2.0 in the run's `default.param`. The ensemble runs
`r result$control$n_particles` particles through
`r result$control$n_iterations` tempered ensemble Kalman updates
(seed `r result$control$seed`).

# Does the calibrated model reproduce the observed target?

```{r}
#| label: fitted-slots
#| fig-cap: "Fitted target slots before (left) and after (right) calibration: the control period mean level and the seven system contrasts. Blue points are the ensemble members, the black dash their mean, red the observation with one standard deviation."
#| fig-width: 10
#| fig-height: 4.5
sm <- result$obs$meta
library(patchwork)
p1 <- plot_target_slots(G_list[[1]][, sm$slot, drop = FALSE], sm,
                        ylab = unique(sm$units), title = "before calibration")
p2 <- plot_target_slots(G_list[[length(G_list)]][, sm$slot, drop = FALSE], sm,
                        ylab = unique(sm$units), title = "after calibration")
p1 + p2
```

```{r}
#| label: scores
soc <- scores[scores$variable == "SOC_periodmean_salinas", ]
kable(soc[, c("iteration", "rmse", "bias", "coverage", "crps")], digits = 2,
      caption = "Fit of the SOC target by tempering step. Iteration 1 is the prior forward run; the last row is the posterior.")
```

The prior ensemble misses the target with an RMSE of
`r round(soc$rmse[1], 1)` `r unique(sm$units)`; the calibrated ensemble closes
that to `r round(soc$rmse[nrow(soc)], 1)` with coverage
`r round(soc$coverage[nrow(soc)], 2)` of observations inside the 90 percent
band.

# What did the data constrain?

```{r}
#| label: shift-table
kable(shift[, c("param", "prior_mean", "post_mean", "post_sd", "var_reduction")],
      digits = 3,
      caption = "Prior to posterior shift per parameter. var_reduction is the fractional loss of prior variance; a value near zero means the data did not inform the parameter.")
```

```{r}
#| label: som-density
#| fig-cap: "Prior and posterior for som_respiration_rate, the one parameter the target strongly constrains."
#| fig-width: 6
#| fig-height: 3
d <- rbind(
  data.frame(stage = "prior",     value = result$U0[, "som_respiration_rate"]),
  data.frame(stage = "posterior", value = result$U[, "som_respiration_rate"])
)
ggplot(d, aes(value, color = stage)) +
  geom_density(linewidth = 1) +
  labs(x = "som_respiration_rate (g C g-1 soil C yr-1)", y = "density", color = NULL)
```

```{r}
#| label: som-numbers
#| include: false
sr <- result$U[, "som_respiration_rate"]
srq <- quantile(sr, c(0.025, 0.975))
vr <- shift$var_reduction[shift$param == "som_respiration_rate"]
```

The target constrains one process rate: `som_respiration_rate` collapses from
its uniform prior to `r round(mean(sr), 3)`
[`r round(srq[1], 3)`, `r round(srq[2], 3)`], a
`r round(100 * vr)` percent variance reduction, and it does so in the first
tempering step (below). `turn_over_time` and `fracLitterRespired` barely
narrow: the period mean SOC target carries little information about how carbon is
partitioned on its way to the soil pool, only about how fast the pool turns
over. The initial stocks stay at their observation-anchored priors, which is
what the anchoring is for.

```{r}
#| label: trace
#| fig-cap: "Ensemble mean and 5-95 percent range of each parameter across tempering steps; dashed lines are the declared support bounds. Initial stocks are in kg C m-2, in model units."
#| fig-width: 10
#| fig-height: 6
pm <- result$eki$par_map
trace <- c(list(result$U0), lapply(result$eki$eki_list, function(s) pm$inv(s$U)))
plot_param_trace(trace, prior)
```

# Does the model capture the compost effect?

The curated treatment pair table names systems 1 and 2 as the compost
comparison: the same quadrennial cover crop with and without 7.6 Mg/ha compost
per crop. The effect is computed per ensemble member and year as treatment
minus control, from the raw per year stocks of the prior (itr1) and posterior
(itr4) forward runs.

```{r}
#| label: effect-harvest
#| include: false
em <- obs$meta[obs$meta$variable == "SOC_stock_salinas" &
               obs$meta$treatment_id %in% c("socs_sys1", "socs_sys2"), ]
vm <- config$forward$var_map["SOC_stock_salinas"]
win <- cbind(socs_sys1 = c(2005L, 2011L), socs_sys2 = c(2005L, 2011L))
out <- file.path(config$forward$workspace, "output")
G_prior <- harvest_output_to_G(file.path(out, "itr1", "out"), em, vm, win)
G_post  <- harvest_output_to_G(file.path(out, "itr4", "out"), em, vm, win)
eff <- function(G) mean(colMeans(G[, em$slot[em$treatment_id == "socs_sys2"]]) -
                        colMeans(G[, em$slot[em$treatment_id == "socs_sys1"]]))
meas <- with(em, mean(value[treatment_id == "socs_sys2"] -
                      value[treatment_id == "socs_sys1"]))
```

```{r}
#| label: effect-figure
#| fig-cap: "Measured against modeled compost effect (system 2 minus system 1), before and after calibration, as the absolute effect and as percent of control. Bands are the 5-95 percent ensemble range; red points the measured contrast with one standard deviation."
#| fig-width: 9
#| fig-height: 4
plot_treatment_effect(G_prior, G_post, em, "SOC_stock_salinas",
                      treatment = "socs_sys2", control = "socs_sys1")
```

Averaged over the period, the measured effect is
`r round(meas, 1)` Mg C/ha; the prior ensemble puts it at
`r round(eff(G_prior), 1)` and the calibrated ensemble at
`r round(eff(G_post), 1)`. The calibration was fit to the contrasts, so
matching their mean is expected; that the posterior band also tracks the
year to year shape in both the absolute and relative form is the substantive
check.

# The held-out N2O contrast

```{r}
#| label: validation
#| fig-cap: "Held-out validation: predicted against measured Modesto N2O treatment contrast per measurement date, with the 5-95 percent ensemble range and a 1:1 line. Axes are linear because a contrast is signed."
#| fig-width: 8
#| fig-height: 4
vmeta <- result$obs_all$meta
vmeta <- vmeta[vmeta$variable == "N2O_contrast_modesto", , drop = FALSE]
plot_validation(result$G_validation, vmeta)
```

```{r}
#| label: val-scores
kable(score_iteration(result$G_validation, vmeta)[, c("variable", "slots", "rmse", "bias", "coverage")],
      digits = 2, caption = "Scores of the held-out contrast at the posterior.")
```

The model predicts a positive compost minus control N2O contrast on most dates
while several measured contrasts are negative. That sign structure is why the
variable is held out: a target the model cannot reach cannot be calibrated
through; the estimator would answer by pushing a nitrogen parameter to its
bound for reasons unrelated to the process it stands for. Held out, the
mismatch stays visible and scored instead of being absorbed.

# Limitations

- The compost carbon inputs carry an amendment quality scaling sourced from the
  literature, not fitted; its uncertainty is not propagated here.
- One process rate dominates the fit. With a single site's period mean and
  contrasts, `turn_over_time` and `fracLitterRespired` are close to
  unidentifiable, and their posteriors should be read as priors.
- The SOC inference rests on one trial; transfer to other soils and systems is
  an assumption until a second site joins the target.
- The establishment years (2003 to 2004) are excluded by design: the first year
  drop is a one time disturbance a first order soil model should not be forced
  to reproduce.
- The N2O response has the wrong sign structure at Modesto; nitrogen parameters
  were not calibrated and the held-out scores quantify, not fix, that gap.

# Reproducing this report

The run is reproduced by the script sequence in this directory's README; this
document then renders from the run's caches with
`quarto render examples/2_joint_soc_n2o/report.qmd`. The only computation here
beyond reading caches is the compost effect harvest, which reads the stored
model output of the first and last iteration.
